---
title: "Accessible Plant Identification for People with Visual Impairment"
author: "John M. A. Wojahn, Patricia M. Wojahn"
date: "2019"
output:
#  bookdown::pdf_document2:
#    toc: TRUE
  bookdown::html_document2: 
    toc: TRUE
    toc_float: TRUE
    theme: flatly
    self_contained: true
link-citations: yes
fontsize: 20pt
urlcolor: blue
csl: AmJBot.csl
bibliography: References.bib
editor_options: 
  markdown: 
    wrap: sentence
---

```{js logo-js, echo=FALSE}
$(document).ready(function() {
});
```

```{r packages, echo=FALSE, warning=FALSE, include=FALSE}
library(knitr)
library(rmarkdown)
library(bookdown)
#library(distill)
library(knitcitations)
library(formatR)
library(devtools)
library(kfigr)
library(dplyr)
library(kableExtra)
library(tufte)

#Generate BibTex citation file for all R packages used to produce report
knitr::write_bib(.packages(), file = 'packages.bib')
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Nervation

(i.e. Leaf Rib Pattern)

Lightly stroke the leaf from bottom to apex, noting when the primary nerve branches.  If it branches immediately from the petiole, count the number of branches (3 for triplinerve vs 5 for palmate). If there is only one nerve, feel for secondary nerves using two fingers: one placed on the main nerve and one tracing the path of a branching nerve from midrib to margin.  Using two fingers allows for easier detection of curvature in the branched nerves (no curvature for parallel vs curvature for penninerve).  If one cannot feel anything other than the center nerve the leaf is reticulate.  

# Composite vs. Simple Leaf

(i.e. One Big Leaf vs Many Small Leaflets on a Stalk)

To identify whether a leaf is simple or composite, run one’s finger down the primary venation of a leaf/leaflet, then down the cylinder to which the leaf/leaflet is attached.  If the leaf is compound, the first cylinder one feel will feel distinctly different from the second cylinder one feel (the branch), and one will also be able to feel the insertion point of the first cylinder into the second as a swelling. To determine type of composite leaf run one’s finger up the rachis to the apex, noting whether or not a terminal leaflet is present (imparipinnate vs paripinnate)

# Phyllotaxy

(i.e. Whether the Leaves Are Across From Each Other Or Alternate on Branch)

Run one’s fingers along a branch and feel the insertion points of each leaf.  Do this for at least 3 leaves (across from each other means opposite whereas not across from each other means alternate).

# Leaf Orientation

(i.e. How Are the Leaves Orientated with Respect to Their Peers)

If using fresh specimens make sure that one does not bed the leaves as one feels them, just gently pinch the the air around the branch of interest, noting the natural orientation of each leaf.  Do this for at least 3 leaves (spiral vs verticillate (in separate whorls) vs planar vs, decussate (in alternating pairs).  To identify leaf orientation using herbarium specimens, one must rely on feeling the petioles for evidence of torsion. 

# Leaf Margin

Gently run one’s finger along the leaf edge to feel for dentition/undulation.  If one feels something other than smooth edges, first note whether the edges feel pointy (some sort of toothedness) or rounded (crenate for small lobes, lobate for larger ones), then draw one’s fingertip from right to left then left to right.  Note if there is a difference in texture (dentate teeth point outward and do not feel different if brushed from different directions vs serrate which has a directionality like a knife).  Next use one’s fingernail to feel for smaller teeth between the teeth (each tooth is a ‘catch’ in the nail’s track) to check for double serration.

# Hairy vs. Glabrous

Caress the axial (top) and abaxial (bottom) surface of each leaf, feeling for hair with one’s most sensitive finger.  Alternatively, one can use the lips to feel the leaf texture.

# Leaf Shape

First feel the primary nerve from base to tip to get a sense of the length of the leaf.  Then, starting from the base and working one’s way up to the apex, trace the leaf shape with the index and thumb.  Ovate leaves are longer than they are wide, in other words they are ovals.  Deltoid leaves are pyramid shaped.  Cordate leaves are heart shaped.  Reniform leaves are kidney shaped.  Round leaves are roughly circular. Elliptic leaves are long ovals.  Linear leaves are much longer than elliptic leaves.  Obdeltoid leaves are oval but with a fat bottom. Obdelliptic leaves are the opposite of obdeltoid leaves: they are oval but with a fatter head.

# Trunk Shape

Start from the base of the tree and feel around its circumference, noting any bulging or buttressing.  If several smaller trunks emanate from the ground towards a central point, then the tree has prop roots.  This method is for in situ specimens, as trunks generally are not preserved as herbarium specimens.

# Latex (fresh only)

For in situ testing make a cut into the flesh of the plant and wait a few seconds.  Using a gloved finger touch the cut area and with one’s thumb.  Press the thumb and forefinger together then slowly pull them apart.  If there is resistance and the feeling of a sticky stringiness between one’s fingers, then latex is present. 

# Branching Pattern

Start from the base and work one’s way up as far as one can reach.  If branching occurs above where one can safely reach use a long bamboo pole to feel where branching occurs.  If a tree is exceptionally tall or is in a forest where most of the branches are high up in the canopy this method may not work.

# Thorns

Whenever feeling an unknown plant for the first time without the presence of a sighted assistant be sure to gently approach it in slow, light caresses, feeling for thorns with one’s least important finger.

# Lenticelles

(i.e. Little Rough/Raised Spots On Bark)

Lenticelles can be felt for by gently caressing the bark with one’s fingertips or fingernails.  One’s lips may also be used to feel for especially small/shallow lenticelles.

# Petal Number

Count the petals from the top in a clockwise direction, keeping the first petal grasped between one’s left forefinger and thumb.

# Placentation

(i.e. Arrangement Of The Plant’s Placentae, for Advanced Users)

If the sample is large and fresh one may be able to cut the ovary open and feel around with one’s fingertips.  If the sample is small or pressed, it may not be possible to identify this character using touch.

# Position of Ovary

(Whether Ovary Swelling is Below, Halfway Through, or Within Flower Cup, for Advanced Users)

Squish the flower to find the flower’s ovary (a bulge).  Once it has been located gently feel the flower whorl: if the base of the whorl is below the swelling, then the flower is hypogynous.  If the whorl attaches to a tube that then attaches to the bulb, the flower is perigynous.  If the whorl is attached to the top of the swelling, the flower is epigynous. 

# Floral Symmetry

(i.e. is the flower symmetrical?)

True floral symmetry is difficult to determine without being able to see the petal coloration, though whether or not a flower is zygomorphic or peloric may be crudely estimated by feeling the general shape of the flower.

© 2019 Wojahn and Wojahn
